---
title: "Pathway Simulation for different Variables and Maximum selection"
author: "Ali Shojaee Bakhtiari"
date: "December 13, 2018"
output: html_document
---
# Load libraries
```{r}
library(energy)
packageVersion("energy")

library(parallel)
library(ggplot2)
```

# Simulation Function

The test values are function set values, unless individually changed.

```{r}
pathway_simulation= function(nsamples = 40,
                             y = ifelse(1:nsamples> nsamples/2,1,0),
                             test = sum(y==1),
                             control= sum(y==0),
                             L=10000,
                             p=100,
                             q=L-p,
                             enrichment_percentage_pathway = 0.1,
                             k = round(p*enrichment_percentage_pathway),
                             enrichment_percentage_background = 0.01,
                             l = round(q*enrichment_percentage_background),
                             ESpathway = 5,
                             ESbackground = 1,
                             expressions = matrix(rnorm(n= nsamples * L) , nrow = nsamples),
                             SPZ=10,
                             numsim1 = 1000,
                             numsim2 = 10000-1, parallel_cores= detectCores()-2,
                             sim_mode = c("max")){

library(energy)
# library(parallel)
library(doParallel)
library(ggplot2)
packageVersion("energy")

# The following script checks for the validity of the simulation histogram. "1-prod((90:80)/(100:90))" for spz and l=10
#mean(sapply(1:1000,function(x) sum(sample(1:100,10)%in%(1:10))!=0))


## Set  ESpathway,ESbackground,p,q, L=p+q, k,l and SPZ, numsim1 ( kinda tricky), numsim2 

## 1) Set matrix for nsamples and L= p+q expressions

# nsamples <- 40
# y <- ifelse(1:nsamples> nsamples/2,1,0)
# test <- sum(y==1)
# control= sum(y==0)
# dy <- dist(y)
# L=10000
# p=100
# q=L-p
# enrichment_percentage_pathway <- 0.1
# k <- round(p*enrichment_percentage_pathway)
# enrichment_percentage_background <- 0.01
# l <- round(q*enrichment_percentage_background)
# ESpathway <- 5
# ESbackground <- 1
# expressions <- matrix(rnorm(n= nsamples * L) , nrow = nsamples)
# SPZ=10
# numsim1 <- 1000
# numsim2 <- 10000-1



##  Set up your pathway of size p with k ( variable) enirched genes and ES1

expressions[y==1,1:k] <- matrix(rnorm(n = test* k,mean = ESpathway), nrow = test)

##  p+1 to L=p+q choose l ( lower) and apply ES2.
expressions[y==1,(p+1):(p+l)] <- matrix(rnorm(n = test* l,mean = ESbackground), nrow = test)


# Actual pathway dcor must be lower than maximum dcor below (dilution)
print( paste("Actual Pathway Dcor = ", dcor(expressions[,1:p],y)))
# Actual maximum dcor has all the enriched
print(paste("Dcor of enriched genes within the pathway = ", dcor(expressions[,1:SPZ],y)))



## 5) on 2, 
#     5.1) choose spz numsim1 times
#     calculate dcor_max over numsim times

pathway_dcor_sim <- NULL
for (ii in 1:numsim1){
    pathway_dcor_sim[ii] <-  dcor(expressions[,sample(1:p, size = SPZ)],y)
}
# pathway_dcor_sim_df<- data.frame(pathway_dcor_sim= pathway_dcor_sim)
# plt <- ggplot(pathway_dcor_sim_df, aes(x=pathway_dcor_sim)) + geom_histogram(binwidth=.05, color="red")+theme_bw()
# print(plt)
pathway_dcor_sim_output <- ifelse(sim_mode=="max",max(pathway_dcor_sim),mean(pathway_dcor_sim))
pathway_dcor_sim_output

## 6) on 4, 
#     6.1) for ii in 1:numsim2

#         6.1.1) choose spz numsim1 times
#         calculate dcor_max(jj) over numsim1 times
# calculate the dcor_max over siulaitons

## Non-parallel

# background_dcor_sim <- NULL
# 
# for (ii in 1:numsim2){
#   print(ii)
#   aux_pathway <- sample(1:L, p) # generate an aux pathway the same size as the actual pathway
#   aux_dcor <- NULL
#   for (jj in 1: numsim1){
#     aux_dcor[jj] <-  dcor(expressions[,sample(aux_pathway, size = SPZ)],y)
#   }
#   background_dcor_sim[ii] <- max(aux_dcor)
#   
# }


cl<-makeCluster(parallel_cores)
registerDoParallel(cl)
background_dcor_sim <- foreach (ii = 1:numsim2, .combine = c, .export = c("L", "p", "expressions","SPZ","y","sim_mode"), .packages = "energy" ) %dopar% {
  # print(ii)
  aux_pathway <- sample(1:L, p) # generate an aux pathway the same size as the actual pathway
  aux_dcor <- NULL
  for (jj in 1: numsim1){
    aux_dcor[jj] <-  dcor(expressions[,sample(aux_pathway, size = SPZ)],y)
  }
  ifelse(sim_mode=="max", max(aux_dcor), mean(aux_dcor))
  
}
stopCluster(cl)



##mean(sapply(1:1000,function(x) sum(sample(1:10000,10)%in%(1:10))!=0))
# background_dcor_sim_df<- data.frame(background_dcor_sim= background_dcor_sim)
# plt <- ggplot(background_dcor_sim_df, aes(x=background_dcor_sim)) + geom_histogram(binwidth=.05, color="red")+theme_bw()
# print(plt)

sim_df <- data.frame(simulation= c(background_dcor_sim,pathway_dcor_sim), type=c(rep("Background", numsim2),rep("Pathway", numsim1)))
plt <- ggplot(sim_df, aes(x=simulation, color=type, fill=type)) + geom_histogram(binwidth=.05)+theme_bw()
print(plt)


## 7 calculate p-value. Re,meber to keep 1 on nominator.
p_value <- (sum(background_dcor_sim>pathway_dcor_sim_output)+1)/(numsim2+1)
return(p_value)
}
```

# Analyzing the effect of Effect size on the simulation
In this test we alter the effect size of the phenotype on the pathway genes from a minimum equal to the random effect to the background to the maximum of choice

```{r}
p_value_variable_effect <- sapply(X = 1:5,  function(x) pathway_simulation(ESpathway = x, numsim1 = 100, numsim2 = 999))
print(p_value_variable_effect)
```
# Analyzing the effect of Effect size on the simulation
In this test we alter the effect size of the phenotype on the pathway genes from a minimum equal to the random effect to the background to the maximum of choice. However, we also increase the effect of the phenotype o the background. The purpose of this test is to test the model when the effect size increases but it also affects the background. 
```{r}
p_value_variable_effect_and_background <- sapply(X = 1:5,  function(x) pathway_simulation(ESpathway = x, ESbackground = x/5, numsim1 = 100, numsim2 = 999))
print(p_value_variable_effect_and_background)
```

# Analyzing the effect of pathway size on the simulation
In this test we alter the size of the pathway and see the effect on the pathway simulation. 

```{r}
 p_value_variable_pathway_size <- sapply(X = seq(50,300,50),  function(x) pathway_simulation(ESpathway = 3, p = x, numsim1 = 100, numsim2 = 999))
```

# Analyzing the effect of surrogate pathway size ()
In this test we alter the size of the pathway surrogate size and see the effect on the pathway simulation.

```{r}
 p_value_variable_pathway_size <- sapply(X = c(1,seq(5,20,5)),  function(x) pathway_simulation(ESpathway = 3, SPZ = x, numsim1 = 100, numsim2 = 999))
```

