---
title: "MPEA Power Analysis"
author: "Ali Shojaee Bakhtiari"
date: "January 7, 2019"
output: html_document
---


# Running the power Scripts
In this step we run and save the Power analysis for each of the 4 scenarios i.e. : 
1) Keeping the effect constant and changing Enrichment levels
2) Keeping enrichment level constant and varying effect level.
3) Changing the pathway size compensation window Sizes
4) Changing the pathway sizes.
The default values ( unless changed for the process): \n
1) 1% background enrichment and 10% pathway enrichment \n
2) Expression size = 10000 \n
3) Pathway size= 100 \n
4) Background effect=1, Pathway effect=10 ( Ratio 10:1) \n
5) Pathway reduction compensator size=10 \n


```{r}
# source(file = "power_analysis_similar_effect_variable_enrichment.R")
# source(file = "power_analysis_variable_effect_similar_enrichment.R")
# source(file = "power_analysis_window_size.R")
# source(file = "power_analysis_variable_pathway_size.R")
```

# Uploading Packages 

```{r}
require(reshape2)
require(ggplot2)
```

# Power Analysis of changing Enrichment percentage inside the pathway while keeping the effect sizes equal
In this Scenario we focus on the effect of changing the enrichment level alone while the effect is similar for pathway and background. When Pathway enrichment = 0.01 t is indistinguishable from the background and power increases as the enrichment level increases.

```{r}
p_value_similar_effect_variable_enrichment <- readRDS(file = "p_value_similar_effect_variable_enrichment_mean.RDS")
enrichment_percentage_pathway= seq(0.01,.1, by=0.01)
power_output <- list()
nsamples <- seq(20,100,20)
for (jj in 1:length(enrichment_percentage_pathway)){
 aux <- sapply(p_value_similar_effect_variable_enrichment, FUN = function(x) x[[jj]]< 0.05  )
 power_output[[jj]] <- rowSums(aux)/dim(aux)[2] 
  
}  
names(power_output) <- paste( "Enrichement=",enrichment_percentage_pathway, sep = "")
power_analysis <- melt(power_output)
power_analysis$Sample_size <- nsamples
names(power_analysis) <- c("Power", "Enrichment","Sample_size")
p <- ggplot(data = power_analysis, aes(x = Sample_size, y = Power, color=Enrichment))+ geom_line(size=2)+theme_bw()+
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), legend.text=element_text(size=10), plot.title = element_text(hjust = 0.5))+xlab("Sample Size")+
  ggtitle("Power analysis: Variable Enrichment Constant Effect")
p
```


# Power Analysis of changing Effect size inside the pathway while keeping the enrichment level constant
In this Scenario we focus on the effect of changin the effect size on MPEA power. To this aim we keep the enrichment percentage level constant between the pathway and the background and only change the effect size. Effect size= 1 is indistinguishable from the background.



```{r}
rm(list=ls())
p_value_variable_effect_similar_enrichment <- readRDS(file = "p_value_variable_effect_similar_enrichment_mean.RDS")
effect_size= seq(1,6, by=1)
power_output <- list()
nsamples <- seq(20,100,20)
for (jj in 1:length(effect_size)){
  aux <- sapply(p_value_variable_effect_similar_enrichment, FUN = function(x) x[[jj]]< 0.05  )
  power_output[[jj]] <- rowSums(aux)/dim(aux)[2] 
  
}  
names(power_output) <- paste( "Effect=",effect_size, sep = "")
power_analysis <- melt(power_output)
power_analysis$Sample_size <- nsamples
names(power_analysis) <- c("Power", "Effect","Sample_size")
p <- ggplot(data = power_analysis, aes(x = Sample_size, y = Power, color=Effect))+ geom_line(size=2)+theme_bw()+
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), legend.text=element_text(size=10), plot.title = element_text(hjust = 0.5))+xlab("Sample Size")+
  ggtitle("Power analysis: Variable Effect ")
p
```

# GSEA power analysis while changing Effect size inside the pathway and keeping the enrichment level constant

```{r}
# rm(list=ls())
nsamples <- seq(20,100,20)
power_output <- readRDS("GSEA_power_similar_enrichment_variable_effect.RDS")
power_analysis <- melt(power_output)
power_analysis$Sample_size <- nsamples
names(power_analysis) <- c("Effect","Power","Sample_size")

p <- ggplot(data = power_analysis, aes(x = Sample_size, y = Power, color=Effect))+ geom_line(size=2)+theme_bw()+
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), legend.text=element_text(size=10), plot.title = element_text(hjust = 0.5))+xlab("Sample Size")+
  ggtitle("Power analysis: Variable Effect ")
p
```




# Power Analysis of changing the Pathway compensator window size
In this scenario, we keep every parameter as constant and only change the the Pathway compensator window size.



```{r}
rm(list=ls())
p_value_variable_window_size <- readRDS(file = "p_value_variable_window_size_mean.RDS")
SPZ= seq(5,20, by=5)
power_output <- list()
nsamples <- seq(20,100,20)
for (jj in 1:length(SPZ)){
  aux <- sapply(p_value_variable_window_size, FUN = function(x) x[[jj]]< 0.05  )
  power_output[[jj]] <- rowSums(aux)/dim(aux)[2] 
  
}  
names(power_output) <- paste( "Window Size=",SPZ, sep = "")
power_analysis <- melt(power_output)
power_analysis$Sample_size <- nsamples
names(power_analysis) <- c("Power", "Window_Size","Sample_size")
p <- ggplot(data = power_analysis, aes(x = Sample_size, y = Power, color=Window_Size))+ geom_line(size=2)+theme_bw()+
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), legend.text=element_text(size=10), plot.title = element_text(hjust = 0.5))+xlab("Sample Size")+
  ggtitle("Power analysis: Variable Window Size ")
p

```

# Power Analysis of changing the Pathway Size
In this scenario, we keep every parameter as constant and only change the the Pathway size.


```{r}
rm(list=ls())
p_value_variable_pathway_size <- readRDS(file = "p_value_variable_pathway_size_mean.RDS")
pathway_size= seq(50,300, by=50)
power_output <- list()
nsamples <- seq(20,100,20)
for (jj in 1:length(pathway_size)){
  aux <- sapply(p_value_variable_pathway_size, FUN = function(x) x[[jj]]< 0.05  )
  power_output[[jj]] <- rowSums(aux)/dim(aux)[2] 
  
}  
names(power_output) <- paste( "Pathway Size=",pathway_size, sep = "")
power_analysis <- melt(power_output)
power_analysis$Sample_size <- nsamples
names(power_analysis) <- c("Power", "Pathway_Size","Sample_size")
p <- ggplot(data = power_analysis, aes(x = Sample_size, y = Power, color=Pathway_Size))+ geom_line(size=2)+theme_bw()+
  theme(axis.text=element_text(size=14), axis.title=element_text(size=14), legend.text=element_text(size=10), plot.title = element_text(hjust = 0.5))+xlab("Sample Size")+
  ggtitle("Power analysis: Variable Pathway Size ")
p
```

