---
title: "Comparing Dcor against GSEA P53 dataset"
author: "Ali Shojaee Bakhtiari"
date: "November 16, 2018"
output: html_document
---

## Loading Library
```{r}
library(knitr)
packageVersion("knitr")
library(energy)
packageVersion("energy")

library(broom)
packageVersion("broom")
library(plyr)
packageVersion("plyr")

library(kableExtra)
packageVersion("kableExtra")



set.seed(0)
```

## Loading the expressions and pathway sets of P53 databse

We use the colapsed symbols to genes expression data for expressions. GSEA uses a chip functoin for connecting the uncollapsed to collapsed data. I am currently figuring out how ti perfprms that step. For now not collapsing serves the purpose.
Genotypes are 33 mutants and 17 wild types.
Pathways are read as done is GSEA.
To remain consistent with GSEA we remove the pathways with sizes below 15 and above 500 as well as pathways that have less than 15 genes present in inside the expression matrix.

```{r}
expressions <- read.csv("./gsea p53/P53_collapsed_symbols_for_R.csv", sep = ",", row.names = 1)
genotypes <- read.csv("./gsea p53/P53_genotype.txt", sep=" ", header = F)
genotypes <- as.vector(unname(ifelse( genotypes=="MUT",1,0)))
gene_names <- rownames(expressions)
expressions <- expressions[-1]
expressions <- t(expressions)
dim(expressions)
pathway <- read.csv("./msigdb_v6.2_files_to_download_locally/msigdb_v6.2_GMTs_R/positional gene sets/c1.all.v6.2.symbols.csv", sep = ",", row.names = 1,na.strings = "", colClasses="character")
pathway <- pathway[,-1]
pathway <- t(pathway)
row.names(pathway) <- NULL
pathway <- as.data.frame(pathway)
pathway=lapply(pathway, function(x) x[!is.na(x)] )
pathway=lapply(pathway, function(x) as.character(x))
pathway_collapsed <- pathway[sapply(pathway,function(x) {(15 <= length(x)) &&(length(x) <= 500)&& sum(colnames(expressions)%in%x)>15})]
min(sapply(pathway_collapsed,length))
```



## Loading the results of running GSEA on P53 dataset

GSEA was separately run and the enrichment results are saved. We upload the results and select the significant findings. 

```{r}
gsea_results <- read.csv("./gsea p53/gsea_p53_enrichment_results6.2.csv", sep = ",", row.names = 1)
rownames(gsea_results) <- tolower(rownames(gsea_results))
gsea_results_significant <- subset(gsea_results, NOM.p.val <0.05 )
```

## Running Dcor and Univariate ttest over the pathways

Expressions concordant with specific pathways are chosen and dcor and univariate t.test are applied on them. The univariate test is adjusted for FDR correction.

```{r}
dcor_results <- list()
univariate_results <- list()
univariate_results_adjusted <- list()
for (ii in 1:length(pathway_collapsed)){
pathway_expressions <- expressions[,colnames(expressions)%in% pathway_collapsed[[ii]]]

 dcor_results[[ii]] <- dcor.test(dist(pathway_expressions),dist(genotypes),R=10000)

pvalues <- apply(X = pathway_expressions, MARGIN = 2, function(x) t.test(x ~ genotypes)$p.value)
univariate_results[[ii]] <- pvalues
univariate_results_adjusted[[ii]]<- p.adjust(pvalues, method = "fdr")
}
names(dcor_results) <- names(univariate_results_adjusted) <- names(univariate_results) <- names( pathway_collapsed)
```


## Adjusting the DCOR results for multiple comparison


```{r}
nominal_dcor_pvalues <- (sapply(dcor_results, function(x) x$p.value))

dcor_significant_pathways_p.value <- sort(nominal_dcor_pvalues[nominal_dcor_pvalues <0.05])
head(dcor_significant_pathways_p.value)

FDR_adjusted_dcor_pathway_q_values <- p.adjust(nominal_dcor_pvalues, method = "fdr")
head(sort(FDR_adjusted_dcor_pathway_q_values)) 
```

## Identifying the pathways deemed significant in both GSEA and Dcor

```{r}
Mutually_identified_in_GSEA_dcor <- intersect(names(dcor_significant_pathways_p.value), rownames(gsea_results_significant))

Mutually_identified_in_GSEA_dcor
```

## Identifying the pathways significant in dcor not in GSEA

```{r}
Significant_in_dcor_not_in_GSEA <- setdiff(names(dcor_significant_pathways_p.value),Mutually_identified_in_GSEA_dcor)
pvalue_of_disparities <- rbind(dcor_significant_pathways_p.value[Significant_in_dcor_not_in_GSEA],gsea_results[Significant_in_dcor_not_in_GSEA,]$NOM.p.val)
rownames(pvalue_of_disparities) <- c("DCOR","GSEA")
pvalue_of_disparities%>%
  kable("html", digits = 4, longtable = TRUE) %>%
  kable_styling(bootstrap_options = "striped", font_size = 12, full_width = F) %>%
  row_spec(0, angle = -0)
```

## Identifying the pathways significant in GSEA not in dcor

```{r}
Significant_in_GSEA_not_in_dcor <- setdiff(rownames(gsea_results_significant),Mutually_identified_in_GSEA_dcor)
pvalue_of_disparities <- rbind(nominal_dcor_pvalues[Significant_in_GSEA_not_in_dcor],gsea_results[Significant_in_GSEA_not_in_dcor,]$NOM.p.val)
rownames(pvalue_of_disparities) <- c("DCOR","GSEA")
pvalue_of_disparities %>%
  kable("html", digits = 4, longtable = TRUE) %>%
  kable_styling(bootstrap_options = "striped", font_size = 12, full_width = F) %>%
  row_spec(0, angle = -0)

```

## Univariate Significance analysis on single pathway gene by gene Significant in Dcor and not in GSEA
```{r}
enrichment_ratio_dcor <- NULL
enrichment_ratio_ttest <- NULL
for ( ii in 1:length(Significant_in_dcor_not_in_GSEA) ){
aux_genes <- intersect(pathway_collapsed[[Significant_in_dcor_not_in_GSEA[[ii]]]],colnames(expressions))
aux_pathway_dcor <- dcor.test(dist(expressions[,aux_genes]),dist(genotypes),R=10000)
aux_individual_genes_in_pathway_dcor <- sapply(aux_genes,function(x) dcor.test(dist(expressions[,x]),dist(genotypes),R=10000)$p.value<0.05 )
aux_enrichment_ratio_dcor <- sum(aux_individual_genes_in_pathway_dcor)/length(aux_individual_genes_in_pathway_dcor)
enrichment_ratio_dcor[ii] <- aux_enrichment_ratio_dcor

aux_individual_genes_in_pathway_ttest <- sapply(aux_genes,function(x) t.test(expressions[,x]~genotypes)$p.value<0.05 )
aux_enrichment_ratio_ttest <- sum(aux_individual_genes_in_pathway_ttest)/length(aux_individual_genes_in_pathway_ttest)
aux_enrichment_ratio_ttest
enrichment_ratio_ttest[ii] <-aux_enrichment_ratio_ttest
}

enrichment_ratio_comparison <- rbind(enrichment_ratio_dcor,enrichment_ratio_ttest)
rownames(enrichment_ratio_comparison) <- c("DCOR","GSEA")
colnames(enrichment_ratio_comparison) <- Significant_in_dcor_not_in_GSEA

enrichment_ratio_comparison%>%
  kable("html", digits = 4, longtable = TRUE) %>%
  kable_styling(bootstrap_options = "striped", font_size = 12, full_width = F) %>%
  row_spec(0, angle = -0)
t.test(enrichment_ratio_dcor,enrichment_ratio_ttest)
enrichment_ratios_percenages <- rbind(sum(enrichment_ratio_dcor>enrichment_ratio_ttest)/length(enrichment_ratio_dcor),
sum(enrichment_ratio_dcor<enrichment_ratio_ttest)/length(enrichment_ratio_dcor),
sum(enrichment_ratio_dcor==enrichment_ratio_ttest)/length(enrichment_ratio_dcor))*100
rownames(enrichment_ratios_percenages) <- c ( "More univariate enrichment detected in Dcor", "More univariate enrichment detected in T-test", "Draw" )
colnames(enrichment_ratios_percenages) <- "Percentage"

enrichment_ratios_percenages%>%
  kable("html", digits = 4, longtable = TRUE) %>%
  kable_styling(bootstrap_options = "striped", font_size = 12, full_width = F) %>%
  row_spec(0, angle = -0)

```

## Univariate Significance analysis on single pathway gene by gene Significant in GSEA and not in DCOR
```{r}
enrichment_ratio_dcor <- NULL
enrichment_ratio_ttest <- NULL
for ( ii in 1:length(Significant_in_GSEA_not_in_dcor) ){
aux_genes <- intersect(pathway_collapsed[[Significant_in_GSEA_not_in_dcor[[ii]]]],colnames(expressions))
if ( identical(aux_genes, character(0))) next
aux_pathway_dcor <- dcor.test(dist(expressions[,aux_genes]),dist(genotypes),R=10000)
aux_individual_genes_in_pathway_dcor <- sapply(aux_genes,function(x) dcor.test(dist(expressions[,x]),dist(genotypes),R=10000)$p.value<0.05 )
aux_enrichment_ratio_dcor <- sum(aux_individual_genes_in_pathway_dcor)/length(aux_individual_genes_in_pathway_dcor)
enrichment_ratio_dcor[ii] <- aux_enrichment_ratio_dcor

aux_individual_genes_in_pathway_ttest <- sapply(aux_genes,function(x) t.test(expressions[,x]~genotypes)$p.value<0.05 )
aux_enrichment_ratio_ttest <- sum(aux_individual_genes_in_pathway_ttest)/length(aux_individual_genes_in_pathway_ttest)
aux_enrichment_ratio_ttest
enrichment_ratio_ttest[ii] <-aux_enrichment_ratio_ttest
}


enrichment_ratio_comparison <- rbind(enrichment_ratio_dcor,enrichment_ratio_ttest)
rownames(enrichment_ratio_comparison) <- c("DCOR","GSEA")
colnames(enrichment_ratio_comparison) <- Significant_in_GSEA_not_in_dcor

enrichment_ratio_comparison%>%
  kable("html", digits = 4, longtable = TRUE) %>%
  kable_styling(bootstrap_options = "striped", font_size = 12, full_width = F) %>%
  row_spec(0, angle = -0)
t.test(enrichment_ratio_dcor,enrichment_ratio_ttest)
enrichment_ratios_percenages <- rbind(sum(enrichment_ratio_dcor>enrichment_ratio_ttest,na.rm = T)/length(enrichment_ratio_dcor),
sum(enrichment_ratio_dcor<enrichment_ratio_ttest,na.rm = T)/length(enrichment_ratio_dcor),
sum(enrichment_ratio_dcor==enrichment_ratio_ttest,na.rm = T)/length(enrichment_ratio_dcor))*100
rownames(enrichment_ratios_percenages) <- c ( "More univariate enrichment detected in Dcor", "More univariate enrichment detected in T-test", "Draw" )
colnames(enrichment_ratios_percenages) <- "Percentage"

enrichment_ratios_percenages%>%
  kable("html", digits = 4, longtable = TRUE) %>%
  kable_styling(bootstrap_options = "striped", font_size = 12, full_width = F) %>%
  row_spec(0, angle = -0)

```







