---
title: "Comparing Dcor against MPEA P53 study"
author: "Ali Shojaee Bakhtiari"
date: "December 17, 2018"
output: html_document
---

## Loading Library
```{r}
library(knitr)
packageVersion("knitr")
library(energy)
packageVersion("energy")

library(broom)
packageVersion("broom")
library(plyr)
packageVersion("plyr")

library(kableExtra)
packageVersion("kableExtra")

library(foreach)
packageVersion("foreach")

library(doParallel)
packageVersion("doParallel")

set.seed(0)
```
## Simulating MPEA Surrogate over the array.


```{r}

load(file = "p53_positional.Rdata")
```

```{r}
epsilon=1e-6
nominal_dcor_pvalues <- sapply(1:pathway_number, function(x) (epsilon+(sum(surrogate_dcor[[x]]>dcor_results[x])))/(1+num_simulation))
names(nominal_dcor_pvalues) <- names( pathway_collapsed)
head(sort(nominal_dcor_pvalues),10)
dcor_significant_pathways_p.value <- sort(nominal_dcor_pvalues[nominal_dcor_pvalues <0.05])
length(dcor_significant_pathways_p.value)
nominal_dcor_q_values <- p.adjust(nominal_dcor_pvalues , method = "fdr")
# nominal_dcor_q_values
nominal_dcor_q_values_under_25 <- sort(nominal_dcor_q_values[nominal_dcor_q_values<0.25])
length((nominal_dcor_q_values_under_25))
nominal_dcor_q_values_under_25
```


## GSEA signifcanct pathways p-value and Q-value
```{r}
nrow(gsea_results_significant)

length(rownames(gsea_results_significant[gsea_results_significant$FDR.q.val<0.25,]))
```


## Identifying the pathways deemed significant in both GSEA and MPEA

```{r}
Mutually_identified_in_GSEA_dcor <- intersect(names(dcor_significant_pathways_p.value), rownames(gsea_results_significant))

Mutually_identified_in_GSEA_dcor
```

```{r}
Mutually_identified_in_GSEA_dcor_fdr <- intersect(names(nominal_dcor_q_values_under_25),rownames(gsea_results_significant[gsea_results_significant$FDR.q.val<0.25,]))
length(Mutually_identified_in_GSEA_dcor_fdr)
```


## Identifying the pathways significant in MPEA not in GSEA

```{r}
Significant_in_dcor_not_in_GSEA <- setdiff(names(dcor_significant_pathways_p.value),Mutually_identified_in_GSEA_dcor)
pvalue_of_disparities <- rbind(dcor_significant_pathways_p.value[Significant_in_dcor_not_in_GSEA],gsea_results[Significant_in_dcor_not_in_GSEA,]$NOM.p.val)
rownames(pvalue_of_disparities) <- c("MPEA","GSEA")
pvalue_of_disparities%>%
  kable("html", digits = 4, longtable = TRUE) %>%
  kable_styling(bootstrap_options = "striped", font_size = 12, full_width = F) %>%
  row_spec(0, angle = -0)
```

## Identifying the pathways significant in GSEA not in MPEA

```{r}
Significant_in_GSEA_not_in_dcor <- setdiff(rownames(gsea_results_significant),Mutually_identified_in_GSEA_dcor)
pvalue_of_disparities <- rbind(nominal_dcor_pvalues[Significant_in_GSEA_not_in_dcor],gsea_results[Significant_in_GSEA_not_in_dcor,]$NOM.p.val)
rownames(pvalue_of_disparities) <- c("MPEA","GSEA")
pvalue_of_disparities %>%
  kable("html", digits = 4, longtable = TRUE) %>%
  kable_styling(bootstrap_options = "striped", font_size = 12, full_width = F) %>%
  row_spec(0, angle = -0)

```




# Close cluster
```{r}
# stopCluster(cl)

```

# PCA plot of pathways deemed significant in MPEA but not GSEA

```{r}
library(ggfortify)
for (name in (sort(Significant_in_dcor_not_in_GSEA, decreasing = F))){
strongestpathwaydcor <- unname(unlist(pathway_collapsed[name]))
expressions_pathway <- as.data.frame(expressions[,which(colnames(expressions)%in% strongestpathwaydcor)])
genotypes_category <- ifelse(genotypes==1, "MUT", "WT")
genotypes_category <- as.data.frame(genotypes_category)
print(name)
print(autoplot(prcomp(expressions_pathway), data = genotypes_category, colour='genotypes_category' )+theme_bw()) 

# pcoa.obj <- pcoa(D = dist(expressions_pathway), correction="none", rn=NULL)
# biplot.pcoa(pcoa.obj)
}



```

# PCA plot of pathways deemed significant in GSEA but not MPEA


```{r}
rownames(gsea_results_significant)
for (name in (Significant_in_GSEA_not_in_dcor)){
  strongestpathwaydcor <- unname(unlist(pathway_collapsed[name]))
  expressions_pathway <- as.data.frame(expressions[,which(colnames(expressions)%in% strongestpathwaydcor)])
  genotypes_category <- ifelse(genotypes==1, "MUT", "WT")
  genotypes_category <- as.data.frame(genotypes_category)
  print(name)
  print(autoplot(prcomp(expressions_pathway), data = genotypes_category, colour='genotypes_category' )+theme_bw()+ggtitle(label = "name")) 
  
  # pcoa.obj <- pcoa(D = dist(expressions_pathway), correction="none", rn=NULL)
  # biplot.pcoa(pcoa.obj)
}
```





